The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-07-31 paper.tex
update paper, nearly finished

\begin{table}
  \centering
  \caption{Table Title}
  \begin{tabulary}{\textwidth}{lLLl}
    \toprule
	ANNOdc class					&Examples				&FMA equivalent / ID						&GFO superclass	\\
	\midrule
	\anno{AnatomicalEntity}			&						&Physical anatomical entity \fma{61775}\\
	\anno{AnatomicalStructure}		&						&Material anatomical entity	\fma{67165}		&\gfo{MaterialObject}\\							
	\anno{SpatialAnatomical entity}	&						&immaterial anatomical entity \fma{67112}	&\gfo{SpaceEntity}\\
	\anno{AnatomicalSpace}			&						&											&\gfo{SpaceRegion}\\
	\anno{AnatomicalSurface}			&Median Plane			&Anatomical surface \fma{24137}			&\gfo{SurfaceRegion}\\
	\anno{AnatomicalLine}				&Mandibular Angle (ppam dexter-go dexter-piam dexter), Sagittal Axis&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
	\anno{AnatomicalPoint}			&Gonion					&Anatomical point \fma{9658}							&\gfo{PointRegion}\\
	\anno{AnatomicalPlane}			&Median Plane			&Anatomical plane \fma{242982}\\
	\anno{AnatomicalAxis}		 		&Sagittal Axis\\
	\anno{Skeleton}					&						&Osseous skeleton \fma{23875}\\
	\anno{BoneStructure}\\
	\anno{Bone}						&Mandibula				&Bone organ \fma{5018}\\
	\anno{BonePart}					&Protuberantia mentalis	&Related to Segment of bone organ \fma{281808}, Zone of bone organ \fma{10483}\\
	\anno{BoneCompound}				&Cranium				&Comparable to union of Skeletal system \fma{23881} and subdivision of skeletal system \fma{85544}\\
	\anno{ToothStructure}				&Dentes Permanentes		&Permanent teeth (\fma{75152}\\%; TA2ID:913)\\
	\anno{Tooth}						&Dens caninus			&Tooth \fma{12516}\\
	\anno{ToothPart}					&Cuspis dentis			\\%\fma{56481}; TA2ID:925\\
	\anno{Phenotype}					&Height, Sex			&											&\gfo{Attributive}\\							
	\anno{AnatomicalProperty}			&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape relation	&\gfo{Attributive}\\
	\anno{RelativeAnatomicalLocation}&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&\gfo{Relator}\\
\bottomrule
\end{tabulary}
\end{table}

@@

An anatomical direction or relative anatomical location, as shown in \cref{fig:relativelocation}, is the position of an anatomical entity (e.g., a bone structure) in relation to another anatomical entity (e.g., another bone, an axis, a plane or a point such as the centre of the body).
Relative anatomical locations are classified according to the anatomical terms of location (e.g., cranial or superior = 'towards the end of the skull', dexter = 'right' or distal = 'in direction towards the end of a limb').
For example, the Glabella is located laterally to both Arcus superciliaris (Arcus superciliaris sinister und Arcus superciliaris dexter).
A relative anatomical location can be considered as an individual relation between two anatomical entities (target entity and reference entity) and can therefore be modelled using a GFO relator.
Relators are composed of roles (in our case target and reference) and have the power to relate arbitrary entities (in our case, anatomical entities)~\citep{gfocategory}.

@@

\subsection{Phenotype}
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/phenotype.png}
\caption{
Exemplary assessment of anatomical properties for the derivation of the phenotype sex, where five cranial bone parts are evaluated for their level of morphological expression.
Method after Walker~\cite{datacollection}.
Image based on TODO%\cite{}.
}\label{fig:gfo}
\end{figure}

@@

% ć in Vodanović cannot be displayed without font package and I don't know if the journal allows a different font
As an example, the discriminant function no. 6 developed by Vodanovic uses the measurement of the angus mandibulae to assign sex, thus deriving a phenotype.
For the left side it is given as:
\[
0.17 \cdot [\text{ppam sinistrum-go sinistrum-paim sinistrum}] - 20.43
\]
The right side is calculated analogously using (ppam dexterum-go dexterum-paim dexterum).
%\[
%\text{For the right side: } ([\text{ppam dexterum-go dexterum-paim dexterum}] \times 0.17) - 20.43
%\]
The threshold or sectioning point is given at 0.2, meaning that if the result is greater than this value, then sex is assigned to male if it is lower it's assigned to female.

@@

\section{Development of ANNOds}\label{sec:domain}
While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
The template is based on the structure of ANNOdc, so that the entered data is compliant with it (i.e., the ANNOds classes are subclasses of the ANNOdc classes and properties from ANNOdc are used).
The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
This approach ensures intuitive and unimpeded data input and a valid end result.
The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.
In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.

>>>>

\begin{table}
  \centering
  \caption{Classes of ANNOdc along with interlinks to the FMA and the GFO top-level ontology.}
  \label{tab:core}
  \begin{tabulary}{\textwidth}{lLLl}
    \toprule
	ANNOdc class					&Examples				&FMA equivalent / ID						&GFO superclass	\\
	\midrule
	\anno{AnatomicalEntity}			&						&Physical anatomical entity \fma{61775}\\
	\anno{AnatomicalStructure}		&						&Material anatomical entity	\fma{67165}		&\gfo{MaterialObject}\\							
	\anno{SpatialAnatomical entity}	&						&immaterial anatomical entity \fma{67112}	&\gfo{SpaceEntity}\\
	\anno{AnatomicalSpace}			&						&											&\gfo{SpaceRegion}\\
	\anno{AnatomicalSurface}			&Median Plane			&Anatomical surface \fma{24137}			&\gfo{SurfaceRegion}\\
	\anno{AnatomicalLine}				&Mandibular Angle (ppam dexter-go dexter-piam dexter), Sagittal Axis&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
	\anno{AnatomicalPoint}			&Gonion					&Anatomical point \fma{9658}							&\gfo{PointRegion}\\
	\anno{AnatomicalPlane}			&Median Plane			&Anatomical plane \fma{242982}\\
	\anno{AnatomicalAxis}		 		&Sagittal Axis\\
	\anno{Skeleton}					&						&Osseous skeleton \fma{23875}\\
	\anno{BoneStructure}\\
	\anno{Bone}						&Mandibula				&Bone organ \fma{5018}\\
	\anno{BonePart}					&Protuberantia mentalis	&Related to Segment of bone organ \fma{281808}, Zone of bone organ \fma{10483}\\
	\anno{BoneCompound}				&Cranium				&Comparable to union of Skeletal system \fma{23881} and subdivision of skeletal system \fma{85544}\\
	\anno{ToothStructure}				&Dentes Permanentes		&Permanent teeth (\fma{75152}\\%; TA2ID:913)\\
	\anno{Tooth}						&Dens caninus			&Tooth \fma{12516}\\
	\anno{ToothPart}					&Cuspis dentis			\\%\fma{56481}; TA2ID:925\\
	\anno{Phenotype}					&Height, Sex			&											&\gfo{Attributive}\\							
	\anno{AnatomicalProperty}			&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape &\gfo{Attributive}\\
	\anno{RelativeAnatomicalLocation}&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&\gfo{Relator}\\
\bottomrule
\end{tabulary}
\end{table}

@@

An anatomical direction or relative anatomical location, as shown in \cref{fig:relativelocation}, is the position of an anatomical entity (e.g., a bone structure) in relation to another anatomical entity (e.g., another bone, an axis, a plane or a point such as the centre of the body).
Relative anatomical locations are classified according to the anatomical terms of location (e.g., cranial or superior = \enquote{towards the end of the skull}, dexter = \enquote{right} or distal = \enquote{in direction towards the end of a limb}).
For example, the Glabella is located laterally to both Arcus superciliaris (Arcus superciliaris sinister und Arcus superciliaris dexter).
A relative anatomical location can be considered as an individual relation between two anatomical entities (target entity and reference entity) and can therefore be modelled using a GFO relator.
Relators are composed of roles (in our case target and reference) and have the power to relate arbitrary entities (in our case, anatomical entities)~\citep{gfocategory}.

@@

\subsection{Phenotype}
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/phenotype.png}
\caption{
Exemplary assessment of anatomical properties for the derivation of the phenotype sex, where five cranial bone parts are evaluated for their level of morphological expression.
% yes, walker method is published in buikstra
Method after Walker~\cite{datacollection}.
Image based on \cite{datacollection}.
}\label{fig:gfo}
\end{figure}

@@

% ć in Vodanović cannot be displayed without font package and I don't know if the journal allows a different font
As an example, the discriminant function no. 6 developed by Vodanovic uses the measurement of the angus mandibulae to assign sex, thus deriving a phenotype.
For the left side it is given as:
\[
0.17 \cdot [\text{ppam sinistrum-go sinistrum-paim sinistrum}] - 20.43
\]
The right side is calculated analogously using (ppam dexterum-go dexterum-paim dexterum).
%\[
%\text{For the right side: } ([\text{ppam dexterum-go dexterum-paim dexterum}] \times 0.17) - 20.43
%\]
The threshold or sectioning point is given at 0.2, meaning that if the result is greater than this value, then sex is assigned to male, if it is lower it is assigned to female.

@@

\section{Development of ANNOds}\label{sec:domain}
While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
The template is based on the structure of ANNOdc, so that the entered data is compliant with it: The ANNOds classes are subclasses of the ANNOdc classes (see \cref{tab:core}) and properties (see \cref{fig:core}) from ANNOdc are used.
The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
This approach ensures intuitive and unimpeded data input and a valid end result.
The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.
In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.
